---
title: "`R-INLA` intercept model issue"
author: "Adam Howes"
date: "`r Sys.Date()`"
toc: true
number-sections: true
format: 
  html: 
    code-fold: false
    html-math-method: katex
---

# Background

```{r}
#| message: false
#| warning: false

library(INLA)
library(ggplot2)

n <- 100

mu <- 3.5
sd <- 1

y <- rnorm(n, mu, sd)

ggplot(data.frame(y = y), aes(x = y)) +
  geom_histogram()
```

# No prior

The mean of the intercept parameter is the empirical mean, as expected:

```{r}
fit <- inla(y ~ 1, data = data.frame(y))

mean(y)
fit$summary.fixed
```

# Confirming that `mean` and `prec` do not apply to the intercept

Setting priors using `mean` and `prec` in `control.fixed` does not alter the intercept:

```{r}
fit2 <- inla(
  y ~ 1,
  data = data.frame(y = y),
  control.fixed = list(mean = 3.5, prec = 1 / 0.01^2)
)

fit2$summary.fixed

fit3 <- inla(
  y ~ 1,
  data = data.frame(y = y),
  control.fixed = list(mean = 1000, prec = 1000)
)

fit3$summary.fixed
```

# Issue with `prec.intercept`

The standard deviation is $\sigma$, the variance is $\sigma^2$, and the precision is $\tau = 1 / \sigma^2$.
Therefore high variance corresponds to a low precision.
Low variance corresponds to high precision.

First a model with high standard deviation, and as such high uncertainty, about the intercept parameter:

```{r}
high_sd <- 100

fit4 <- inla(
  y ~ 1,
  data = data.frame(y = y),
  control.fixed = list(mean.intercept = 3.5, prec.intercept = 1 / high_sd^2)
)

fit4$summary.fixed

fit4$marginals.fixed$`(Intercept)` |>
  as.data.frame() |>
  ggplot(aes(x = x, y = y)) +
    geom_line()
```

Now a model with low standard deviation, and as such low uncertainty, about the intercept parameter:

```{r}
low_sd <- 0.01

fit5 <- inla(
  y ~ 1,
  data = data.frame(y = y),
  control.fixed = list(mean.intercept = 3.5, prec.intercept = 1 / low_sd^2)
)

fit5$summary.fixed

fit5$marginals.fixed$`(Intercept)` |>
  as.data.frame() |>
  ggplot(aes(x = x, y = y)) +
    geom_line()
```

What about setting the `mean.intercept` to be large? There is no almost no difference here.

```{r}
fit6 <- inla(
  y ~ 1,
  data = data.frame(y = y),
  control.fixed = list(mean.intercept = 1000000, prec.intercept = 1 / low_sd^2)
)

fit6$summary.fixed

fit6$marginals.fixed$`(Intercept)` |>
  as.data.frame() |>
  ggplot(aes(x = x, y = y)) +
    geom_line()
```

What about large `mean.intercept` with a high standard deviation? Now it's recovering 3.5.

```{r}
fit7 <- inla(
  y ~ 1,
  data = data.frame(y = y),
  control.fixed = list(mean.intercept = 1000000, prec.intercept = 1 / high_sd^2)
)

fit7$summary.fixed

fit7$marginals.fixed$`(Intercept)` |>
  as.data.frame() |>
  ggplot(aes(x = x, y = y)) +
    geom_line()
```

# Computing environment

```{r}
sessionInfo()
```