---
title: "Notes on \"Nowcasting by Bayesian Smoothing: A flexible, generalizable model for real-time epidemic tracking\""
author: "Adam Howes"
date: "`r Sys.Date()`"
toc: true
number-sections: true
format: 
  html: 
    code-fold: false
    html-math-method: katex
---

# Introduction

* Nowcasting by Bayesian smoothing (NobBS)
* Demonstrate on dengue cases in Puerto Rico and influenza-like illnesses in the US

# Materials and methods

* $n_{t, d}$ cases reported for week $t$ with delay $d$
* Likelihood: $n_{t, d} \sim \text{Pois}(\lambda_{t, d})$ or $n_{t, d} \sim \text{NB}(r, p_{t, d})$ where $p_{t, d} = r / (r + \lambda_{t, d})$
* Latent: $\log(\lambda_{t, d}) = \alpha_t + \log(\beta_t)$
  * $\alpha_t$ true epidemiological signal with random walk prior $\alpha_t \sim \text{RW}(1)$
  * $\beta_t$ probability of reporting with delay $d$ specified using Dirichlet prior $\beta_d \sim \text{Dir}(\theta)$
  * i.e. random effects for week and for delay
  * $\lambda_{t, d} = \exp(\alpha_t) \times \beta_t$
  * See general chain ladder model from actuarial science
* Hyperparameters
  * $\tau_\alpha \sim \text{Gamma}(0.01, 0.01)$
  * $r \sim \text{Gamma}(60, 20)$
  
# Implementation

* https://github.com/sarahhbellum/NobBS
* `NobBS` R package
* Uses JAGS i.e. a Gibbs sampler (is Martyn Plummer still maintaining this?)


```{r}
# devtools::install_github("sarahhbellum/NobBS")
library(NobBS)
library(ggplot2)
library(dplyr)

data(denguedat)
head(denguedat)

denguedat <- denguedat |>
  mutate(
    report_delay = report_week - onset_week
  )
```

```{r}
ggplot(denguedat, aes(x = report_delay / 7)) +
  geom_histogram() +
  theme_minimal() +
  labs(x = "Reporting delay (weeks)", y = "Count")
```