# Getting started

Notes on: https://blog.djnavarro.net/posts/2021-11-19_starting-apache-arrow-in-r/

* You don't need to care if you work with small rectangular data
* Care if
  1. You have long data
  2. You might need to work between different languages
* Arrow is a standardised, language-indep format for storing table-like data in-memory, with a columnar format and supporting zero-copy reads without serialisation overhead
  * Same thing regardless of the programming language
  * Table-like: not sure what this means formally but it's like a table
  * In-memory: not about "on disk" or file formats
  * Columnar: represents data by columns (unlike CSV which does row-wise) -- this is more efficient for filtering
  * Zero-copy reads without serialisation overhead: ...
  
Interjection, notes on https://blog.djnavarro.net/posts/2021-11-15_serialisation-with-rds/

* The process which takes an object stored in memory and converts it to a stream of bytes that can be written to a file or transmitted elsewhere
* The in-memory representation of a dataframe is a list of vectors

```{r}
df <- data.frame(x = rnorm(100), y = rnorm(100))
write.csv(df, "df.csv")

charToRaw("x")
charToRaw("x") |> rawToBits()

read_bytes <- function(path, max_bytes = 100) {
  con <- file(path, open = "rb")
  bytes <- readBin(con, what = raw(), n = max_bytes)
  close(con)
  return(bytes)
}

read_bytes("df.csv")
```

* The JSON format (JavaScript Object Notation) https://www.json.org/json-en.html
* To serialise an R object we need to store the metadata (like classes, names, other attributes)

```{r}
saveRDS(df, "df.rds")
read_bytes("df.rds")
```

* Stopped here: https://blog.djnavarro.net/posts/2021-11-15_serialisation-with-rds/#the-serialize-function